---
title: "Lab 12 Homework"
author: "Marina Puffer"
format: pdf
---

# Population Scale Analysis 

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

## Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```
Sample size for each genotype:
```{r}
table(expr$geno)
```

```{r}
median(expr$exp[expr$geno=="A/A"])
median(expr$exp[expr$geno=="A/G"])
median(expr$exp[expr$geno=="G/G"])
```
Median expression value for A/A is 31.25, A/G is 25.06, and G/G is 20.07. 



```{r}
library(ggplot2)
```

## Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?
Make a boxplot
```{r}
ggplot(expr)+aes(x=geno, y=exp)+geom_boxplot()
```
Median expression levels for the genotypes are shown above. The A/A allele is more highly expressed than G/G, so the G/G SNP likely decreases the expression of this gene. 

