---
title: "Class09: Halloween Mini-Project"
author: "Marina Puffer(A16341339)"
format: pdf
---

# 1.Importing candy data:

```{r}
candy.file <- "candy-data.txt"
candy=read.csv(candy.file, row.names = 1)
head(candy)
```

## Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

85 different types of candy are in the dataset

## Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

38 fruity types of candy are in the dataset

# 2. What is your favorite candy?

## Q3. What is your favorite candy in the dataset and what is it's `winpercent` value?

My favorite candy is Twix:

```{r}
candy["Twix",]$winpercent
```

Win percent of Twix is 81%

## Q4. What is the `winpercent` value for "Kitkat"?

```{r}
candy["Kit Kat",]$winpercent
```

## Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library("skimr")
skim(candy)
```

The winpercent variable looks to be on a different scale than the others, all of the statistics in that row are much greater than the rest.

## Q7. What do you think a zero and one represent for the `candy$chocolate` column?

A zero means that that particular candy does not contain chocolate, and a 1 means that the candy does contain chocolate.

## Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

## Q9. Is the distribution of winpercent values symmetrical?

No, it is skewed to the right.

## Q10. Is the center of the distribution above or below 50%?

The center is below 50%

## Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
chocolate.rank <-candy$winpercent[as.logical(candy$chocolate)]
mean(chocolate.rank)
fruity.rank <- candy$winpercent[as.logical(candy$fruity)]
mean(fruity.rank)
```
On average, the chocolate candy is higher ranked than fruit candy. 

## Q12. Is this difference statistically significant?
```{r}
t.test(chocolate.rank, fruity.rank)
```
Since the p-value is below 0.05 , the difference between the rankings of fruity and chocolate candy are statistically significant. 

# 3. Overall candy rankings 

## Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters are the least liked candy types.

## Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
tail(candy[order(candy$winpercent),], n=5)
```
Snickers, Kit Kat, Twix, Reese's Minis, and Reese's Peanut Butter cup are the most favorite. 

## Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)
ggplot(candy, aes(winpercent, rownames(candy)))+
  geom_col()
```
## Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy, aes(winpercent, reorder(rownames(candy),winpercent)))+
  geom_col()
```

```{r}
#select column by candy$___, set as.logical then designate color
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols[as.logical(candy$hard)]="grey"
#ggplot using set colors
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

## Q17. What is the worst ranked chocolate candy?
Sixlets

## Q18. What is the best ranked fruity candy?
Starburst

# 4. Taking a look at pricepoint 

## Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
```{r}
library(ggrepel)

# plot of price vs winpercent
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 52)
```
Reese's Miniatures are the best bang for your buck.


## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The top 5 most expensive candies are Nik L Nip, Nestle Smartiesm Ring popm Hershey's Krackel, and Hershey's Milk Chocolate. Of these, Nik L Nip is the least popular. 

## Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment(). 
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

# 5. Exploring the correlation structure
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Chocolate and fruity are anti-correlated.

## Q23. Similarly, what two variables are most positively correlated?
Chocolate and winpercent are the most positively correlated. 

# 6. PCA
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2]) #plot of main PCA score of PC1 vs PC2
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
#add color and change plotting character
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 52)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
#library(plotly)
```
```{r}
#ggplotly(p) 
#plotly code does not render to pdf, only HTML
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. This makes sense because these characteristics show negative correlations with the other variables. On the corrplot, fruity is highly negatively correlated with chocolate and is moderately negatively correlated with several othr variables, such as bar and price. Pluribus is highly negatively correlated with bar and is weakly negatively correlated with most other variables. Hard does not show any negative correlations of high magnitude, but shows small negative coorelations with most other categories in the plot. PCA shows which variables have the highest variance, so this plot shows the variables which have the most negative correlations by placing them on opposite ends of the scale. 
